The effect of clumpy outflows and disc anisotropy on quasar unification scenarios

### 2015-07-20 paper_mnras.tex
added more text

%% declare any packages used
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{color}
\usepackage{pdfpages}
\usepackage{appendix}
\usepackage{subfigure}
%\usepackage{epsfig}
%\usepackage[dvips]{color}
%\usepackage{aabib}

@@

%\author[Matthews et al.]
\author[Matthews et al.]{
\parbox[t]{\textwidth}{
J.~H.~Matthews$^1$\thanks{jm8g08@soton.ac.uk}, C.~Knigge$^1$,
N.~Higginbottom$^1$, K.~S.~Long$^2$, S.~A.~Sim$^3$, and S.~W.~Mangham$^1$
}
\medskip  
\\$^1$School of Physics and Astronomy, University of Southampton, Highfield, Southampton, SO17 1BJ, United Kingdom
\\$^2$Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD, 21218
\\$^3$School of Mathematics and Physics, Queens University Belfast, University Road, Belfast, BT7 1NN, Northern Ireland, UK
}

@@

The spectral energy distributions (SEDs) of 
quasars and luminous active galactic nuclei (AGN) 
typically exhibit a series of strong emission lines (e.g. \la, \civ, \nv) 
with an underlying blue continuum - the so-called {\sl `big blue bump'} (BBB). 
The BBB is normally attributed to emission from a geometrically thin, 
optically thick accretion disc surrounding the central black hole (REF), similar
to that described by \cite{shakurasunyaev1973}.
In addition to the inflowing accreting material, 
outflows are ubiquitous in AGN
and quasars (REFs). These outflows can take the form of 
highly collimated radio jets \citep{bellonijet2010}(BETTER REF), 
or mass-loaded `winds' emanating from the accretion disc. 
Approximately $20\%$ of quasars exhibit broad absorption lines (BALs) in the ultraviolet,
providing clear evidence for outflowing absorbing material
\citep{weymann1991, reichard2003, knigge2008, turnermiller2009, allen2011}.
The simplest explanation for the incidence of 
BAL quasars (BALQSOs) is in terms of an accretion disc wind (ADW). 
Within this paradigm, the BALQSO fraction is associated with
the covering factor of the outflow.

It is possible that the  diverse phenemonology
of luminous AGN and QSOs can be broadly explained by
into a simple picture of geometric unification \citep[e.g.][]{MCGV95, elvis2000}. 
In such a model, a biconical wind rises from 
the accretion disc, and the class of object is explained by the material
intercepting the line of sight. Depending on viewing angle, an observer 
may then see a BALQSO, LoBAL-QSO or normal `Type 1' quasar.
Within this framework, the broad-line region (BLR) is normally
assumed to correspond to the dense wind base.
% As well as acting as a source of photoionized plasma, a 
A disc wind may also have a profound effect on the structure and 
emergent continuum of the accretion disc itself.
Mass-loss will alter the accretion rate and resultant 
temperature of the accretion disc, possibly explaining some 
of the features we typically see in luminous AGN \citep{laordavis2014}.
Recent results from \cite{capellupo2015} find 
that if one includes a combination of mass-loss, general relativity (GR) and Comptonisation
then AGN SEDs can, in general, be fitted well with accretion disc models.
In general, mass-loss treatments appear to be critical if an $\alpha$-disc
model is to successfully fit AGN SEDs, particular in the UV region of the spectrum.

Despite the clear importance of ADWs in understanding AGN SEDs and accretion physics, 
much of the underlying outflow physics remains highly uncertain. 
Several possible driving mechanisms for ADWs have been proposed, including
thermal pressure (REFs), magnetocentrifugal forces \citep{blandfordpayne} and 
radiation pressure on spectral lines \citep[`line-driving'][]{lucysolomon1970, MCGV95}.
Of these, line-driving is possibly the most attractive, due 
to the strong absorption lines seen in BALQSOs and the X-ray spectra of AGN (REFs).
The presence of line-locked features \citep{bowler2014} and the so-called `ghost of \la' \citep{arav1996}
in the spectra of such objects also gives clearer evidence that line-driving is
at least partially contributing to the acceleration of the wind.
The efficiency of line-driving is crucially dependent on the ionization state 
of the outflowing plasma. \cite{MCGV95} proposed a potential solution: 
a region of `hitchhiking gas' that could shield the wind from the central X-ray source. 
Hydrodynamic simulations of line-driven disc winds also found a shielding region
was required to maintain the correct ionization state \citep{PSK2000,PK04}. 
However, \cite{H14} showed that including multiple scattering means the ionizing radiation 
field could still reach the previously shielded regions in those particular models.

An alternative solution is that the wind is clumped (REFs), possibly on multiple scale lengths.
Local density enhancements could lower the ionization parameter of the plasma
while still maintaining the same mass-loss rate and column density. 
An inhomogenous outflow has been proposed as a model for the BLR (REFs)
Clumping is expected in outflows and BLRs.

@@

Here we describe the results from our model.

@@

@@

Present a spectrum of the UV and possible optical too.

>>>>

%% declare any packages used
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{color}
\usepackage{pdfpages}
\usepackage{appendix}
\usepackage{subfigure}
\usepackage{url}
%\usepackage{epsfig}
%\usepackage[dvips]{color}
%\usepackage{aabib}

@@

%\author[Matthews et al.]
\author[Matthews et al.]{
\parbox[t]{\textwidth}{
J.~H.~Matthews$^1$\thanks{jm8g08@soton.ac.uk}, C.~Knigge$^1$,
N.~Higginbottom$^1$, K.~S.~Long$^2$, S.~A.~Sim$^3$ and S.~W.~Mangham$^1$
}
\medskip  
\\$^1$School of Physics and Astronomy, University of Southampton, Highfield, Southampton, SO17 1BJ, United Kingdom
\\$^2$Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD, 21218
\\$^3$School of Mathematics and Physics, Queens University Belfast, University Road, Belfast, BT7 1NN, Northern Ireland, UK
}

@@

The spectral energy distributions (SEDs) of 
quasars and luminous active galactic nuclei (AGN) 
typically exhibit a series of strong emission lines
with an underlying blue continuum - the so-called {\sl `big blue bump'} (BBB). 
The BBB is normally attributed to emission from a geometrically thin, 
optically thick accretion disc surrounding the central black hole (REF), similar
to that described by \cite{shakurasunyaev1973}.
In addition to the inflowing accreting material, 
outflows are ubiquitous in AGN
and quasars \citep{kellerman1989,ganguly2008}. These outflows can take the form of 
highly collimated radio jets \citep[e.g.][]{hazard1963,potash1980,perley1984,marscher2006}, 
or mass-loaded `winds' emanating from the accretion disc 
\citep{weymann1991,turnermiller2009}. 
Outflows in AGN offer a 
natural feedback mechanism through which the central source can 
affect its environment \citep{king2003,king2005,fabian2012}
-- feedback that is required by models of galaxy evolution (REFs) 
and may explain the `$M-\sigma$' relation \citep{silkrees1998,haring2004}.

Approximately $20\%$ of quasars exhibit broad absorption lines (BALs) in the ultraviolet,
providing clear evidence for outflowing absorbing material
\citep{weymann1991, reichard2003, knigge2008, turnermiller2009, allen2011}.
The simplest explanation for the incidence of 
BAL quasars (BALQSOs) is in terms of an accretion disc wind (ADW). 
Within this paradigm, the BALQSO fraction is associated with
the covering factor of the outflow.
In addition, ADWs may offer a natural explanation for the
diverse phenemonology of luminous AGN and QSOs \citep[e.g.][]{MCGV95, elvis2000}. 
In such a model, a biconical wind rises from 
the accretion disc, and the class of object is explained by the material
intercepting the line of sight. Depending on viewing angle, an observer 
may then see a BALQSO, LoBAL-QSO or normal `Type 1' quasar.
Within this framework, the broad-line region (BLR) is normally
assumed to correspond to the dense wind base.
% As well as acting as a source of photoionized plasma, a 
A disc wind may also have a profound effect on the structure and 
emergent continuum of the accretion disc itself.
Mass-loss will alter the accretion rate and resultant 
temperature of the accretion disc, possibly explaining some 
of the features we typically see in luminous AGN \citep{laordavis2014}.
Recent results from \cite{capellupo2015} find 
that if one includes a combination of mass-loss, general relativity (GR) and Comptonisation
then AGN SEDs can, in general, be fitted well with accretion disc models.
Mass-loss therefore appears to be critical if an $\alpha$-disc
model is to successfully fit AGN SEDs, particular in the UV region of the spectrum.

Despite the clear importance of ADWs in understanding AGN SEDs and accretion physics, 
much of the underlying outflow physics remains highly uncertain. 
Several possible driving mechanisms for ADWs have been proposed, including
thermal pressure \citep{weymann1982, begelman1991}, magnetocentrifugal forces 
\citep{blandfordpayne,pelletier_pudritz} and 
radiation pressure on spectral lines \citep[`line-driving';][]{lucysolomon1970,shlosman1985,MCGV95}.
Of these, line-driving is possibly the most attractive, as
strong absorption lines are already seen in BALQSOs and the X-ray spectra of AGN 
\citep{reeves2003,poundsreeves2009,tombesi2010a}.
The presence of line-locked features \citep{bowler2014} 
and the `ghost of \la' (Arav et al. 1996; Arav 1996; North 2006; but see 
also Cottis et al. 2010) \nocite{arav1995, arav1996, north2006,cottis2010}
in the spectra of some BALQSOs also gives clearer evidence that line-driving is
at least partially contributing to the acceleration of the wind.

The efficiency of line-driving is crucially dependent on the ionization state 
of the outflowing plasma. \cite{MCGV95} proposed a potential solution: 
a region of `hitchhiking gas' that could shield the wind from the central X-ray source. 
Hydrodynamic simulations of line-driven disc winds also found a shielding region
was required to maintain the correct ionization state \citep{PSK2000,PK04}. 
However, \cite{H14} showed that including multiple scattering means the ionizing radiation 
field could still reach the previously shielded regions in those particular models.

An alternative solution is that the wind is clumped (REFs), possibly on multiple scale lengths.
Local density enhancements could lower the ionization parameter of the plasma
while still maintaining the same mass-loss rate and column density. 
Clumping is expected in outflows and the BLR in a number of separate models
with different motivations.

Physically, clumping can be present due to magnetic confinement of 
dense regions (REFs) or the instabilities that are theoretically predicted 
in line-driven winds (REFs).

% An inhomogenous outflow has been proposed as a model for the BLR (REFs)
% Clumping is expected in outflows and BLRs.

@@

Here we describe the results from our model. This set of parameters
was arrived at by conducting a limited grid search over a 
5-dimensional parameter space involving the variables
$r_{min}$, $\theta_{min}$, $f$, $\alpha$ and $R_V$.
We then evaluated these models qualitavely based on the following
criteria:

\begin{itemize}
\item Does the model maintain the correct ionization state and produce strong BALs?
\item Does strong line emission emerge at low inclinations?
\item Do H recombination lines appear in the spectrum?
\item Do a certain range of angles produce LoBAL features?
\end{itemize}

We arrived at a model which is one of the most promising,
but also representative of a family of models. Examination 
of this individual model allows us to gain insight
into fundamental geometrical and physical constraints, 
as discussed in section 5. 
The full grid, including output spectral files and plots can be found at
\url{jhmatthews.github.io/agn-grid-archive}.

@@

\begin{table}
\begin{tabular}{p{3cm}p{4cm}}
\hline Free Parameters 	&	 Value \\ 
\hline \hline 
$M_{BH}$ 	 &	 $1\times 10^9~\rm{M_{\odot}}$ \\ 
$\dot{M}_{acc}$ 	 &	 $5~M_{\odot}yr^{-1} \simeq 0.2~\dot{M}_{Edd}$\\ 
$\alpha_X$ 	 &	 $-0.9$ \\ 
$L_{X} $ 	 &	 $10^{45}~\rm{ergs~s^{-1}}$\\ 
$r_{disc}(min)=r_{X}$   &	 $6r_g=8.8\times10^{14}~{\rm cm}$ \\ 
$r_{disc}(max)$   &	 $3400r_g = 5\times10^{17}~{\rm cm}$ \\ 
$\dot{M}_{wind}$  &	 $5~M_{\odot}yr^{-1}$ \\ 
$r_{min}$ 	&	 $300r_{g} = 4.4\times10^{16}~{\rm cm}$\\ 
$r_{max}$ 	&	 $600r_{g} = 8.8\times10^{16}~{\rm cm}$ \\ 
$\theta_{min}$ 	&	 $70.0^{\circ}$ \\ 
$\theta_{max}$ 	&	 $82.0^{\circ}$ \\ 
$\lambda$ 	&	 $0$ \\ 
$v_{\infty}$ 	&	 $v_{esc}$(f=1) \\ 
$R_v$ 	        &	 $10^{19}$cm \\ 
$\alpha$ 	&	 $1.0$ \\
$f$ 	&	 $0.01$ \\
\hline 
% Derived Parameters 	&	 Value \\ 
% \hline \hline
% $L_{\nu}(2500\mbox{\scriptsize{\AA}})$&	 $6.3\times10^{30}~\rm{ergs~s^{-1}~Hz^{-1}}$\\ 
% $L_{\nu}(2keV)$   &	 $1.2\times10^{25}~\rm{ergs~s^{-1}~Hz^{-1}}$\\ 
% $L_{bol}$ 	 &	 $2.4\times10^{46}~\rm{ergs~s^{-1}}$\\
% $M_{bol}$ 	 &	 -27.4\\ 
% $M_u$            &	 -26.2\\ 
%  $\alpha_{OX} $ 	 &	 -2.2\\ 
\end{tabular}
\caption{Wind geometry parameters used in the model.}
\label{wind_param}
\end{table}

@@

Figure 4 shows the synthetic spectrum in the UV from the two models, with the
benchmark model from H13 also shown for comparison. We also show a comparison
to composite quasar and BALQSO spectra in figure 5, with the optical waveband 
included. In the UV, strong BAL profiles are formed in both models. This
highlights the first success of our model: clumping allows the correct ionization state 
to be maintained in the presence of strong X-rays. 
The BAL profiles are saturated, 
and generally show the strongest absorption at the terminal velocity.
This contradicts observed profiles of many BALQSOs (REFs), 
although there are some objects which do show this behaviour (REFs). 
However, we have found that this profile shape is by no means ubiquitous.
Figure 6 shows how the line profile varies with angle for three of the UV lines
(see also section 4.4). In addition, the line profile
shape is strongly dependent on the density, ionization and velocity 
profiles intersected by the line of sight. Thus, small tweaks of the velocity
law and angular distributions of streamlines can dramatically alter
the line profile shape. Figure ?? shows selection of line profiles 
for different kinematic parameters to demonstrate this. 
The degeneracy between parameters, and lack of available
physical and observational constraints goes someway to showing why individual 
fitting of line profile shapes with kinematic models is such a difficult
exercise.

We find that model A can produce significant line emission
at low inclinations, particular in \civ, 
and the improved treatment of recombination results in a strong \la\
line. However, the red wing of the BAL profiles 
is far stronger than seen in BALQSO spectra and composites.
This is due to a combination of strong collisionally excited emission
and continuum attenuation. To illustrate the amount of 
continuum attenuation we also show the continuum from a disc-only model.
The line-to-continuum ratios at low inclinations are also
significantly affected by disc foreshortening and limb darkening.
The angular distribution of the disc radiation is clearly
crucially important in determining the emergent line equivalent widths 
(EWs) -- this is discussed further in section~5.

To assess the ability of the model to match real 
quasar spectra, we show a comparison to an HST quasar composite in figure 5. 
All the spectra are normalised to the flux at 2000\AA in these plots.
Model A matches the \civ\ line profile well, but the \la\ profile
is a factor of $\sim3$ weaker than in the composite spectrum.
Figure 5 also shows the spectrum into the optical,
showing that the outflow produces signiciant optical emission lines
as well as having a noticeable effect on the continuum shape. This
is particularly true around the Balmer jump at $3646\AA$ and is due to
a prominent wind-formed recombination continuum.
